Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-17 analysis/paper/paper.Rmd
work on GM background, and add the metric analysis back

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015; @Selden2019].

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes. EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984]. Most applications are associated with human remains and animal bones to study biological variation, and has been increasing for stone artifacts to explore stone tool topology and technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Despite the EFA applications for ceramics is rare, they demonstrate the potential of this promising approach for answering questions related to ceramic taxonomy [@Topi2018; @Wilczek2014]. For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.

@@

The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot)). PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare in the neck, from a curved to straight shape. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

@@

>>>>

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons[@Birch2019; @Buchanan2019; @Cardillo2010; @Lycett2013; @Selden2018], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].

Geometric morphometrics methods can provide a useful framework for further statistical evaluation on shape variability of archaeological assemblages to explore how they change over time or across geographic areas to answer questions related to variation in morphology and standardization. A common procedure is using multivariate statistical methods, such as principal component analysis or canonical variate analysis, to capture the important variances of the overall variation in shape variables generated by GM and then visualize the variances to identify any clustered analytical groups that suggest less variability or high standardization. For example, the application of GM to stone tools coupled with statistical tests, such as coefficient of variation on principal component scores [@Archer2015], and on a series of metric variables including length, width, or thickness [@Buchanan2018; @Okumura2014; @Smith2016], or Kruskal-Wallis test on flake scar patterns [@Eren2015], which allows further discussion of causes of standardization and variability that may relate to degree of tool reduction, specialized function, or lack of innovation in a small group. In addition, cluster analysis such as k-mean aggregation for shape variables generated by GM is another way to discuss the variability among shape groups [@Doyon2019].

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes. EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984]. Most applications are associated with human remains and animal bones to study biological variation, and has been increasing for stone artifacts to explore stone tool topology and technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Despite the EFA applications for ceramics is rare, this approach is promising for answering questions related to ceramic taxonomy [@Topi2017; @Wilczek2014]. For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.

@@

The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principal component. With the third component, the first three principal components explain `r vari_three`% of the total variance. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot)). PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare in the neck, from a curved to straight shape. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

@@

We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)). Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence. After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05). Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables. In addition, the result of spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery shape samples presents multiple clusters with high density of pottery during the European presence. However, the hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence. In contrast, the distribution of pottery is more similar to the random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. Or, this may indicate that there are other factors lead to pottery shape standardization instead of a small number of artisans.

```{r metricCVstest}
# select metric attributes and tidy data for computing CV
kwl_p_metrics_long <-
  kwl_p_ave %>%
  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
  select(period, 
         Dia_Rim_ave, Dia_Neck_ave, 
         Dia_Body_ave, Thick_Rim_ave, 
         Thick_Neck_ave, Thick_Body_ave,
         dia_rim_body, thi_rim_body,) %>%
  rename(`Rim diameter (mm)` = Dia_Rim_ave,
         `Neck diameter (mm)` = Dia_Neck_ave,
         `Body diameter (mm)` = Dia_Body_ave,
         `Rim thickness (mm)` = Thick_Rim_ave,
         `Neck thickness (mm)` = Thick_Neck_ave,
         `Body thickness (mm)` = Thick_Body_ave,
         `Ratio of Rim/Body diameter` = dia_rim_body,
         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
  gather(variable, value, -period) %>%
  filter(!is.na(period), !is.na(value)) %>% 
  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
                                                "Neck thickness (mm)",
                                                "Body thickness (mm)",
                                                "Ratio of Rim/Body thickness",
                                                "Rim diameter (mm)",
                                                "Neck diameter (mm)",
                                                "Body diameter (mm)",
                                                "Ratio of Rim/Body diameter"))) %>%
  arrange(variable)
# filter pre-e and post-e periods
kwl_p_metrics_long_pre_post <-
  kwl_p_metrics_long %>%
  filter(!period == "ch-con")
# filter post-e and chi periods
kwl_p_metrics_long_post_chi <-
  kwl_p_metrics_long %>%
  filter(!period == "pre-e")
#---------------------------CV test for pre-e and post-e pair
kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
  kwl_p_metrics_long_pre_post %>%
  nest(-variable) %>%  # or nest(period, value) %>%
  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
                               ~bind_cols(asymptotic_test (
                                 .x$value,
                                 .x$period)))) %>%
  unnest(asymptotic_test)
# compute CV
kwl_p_cvs_pre_post <-
  kwl_p_metrics_long_pre_post %>%
  group_by(period, variable) %>%
  summarise(cvs = raster::cv(value, na.rm = TRUE))
# join p-values to CVs
kwl_p_cvs_and_pvalues_pre_post <-
  kwl_p_cvs_pre_post %>%
  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
  mutate(variable = factor(variable, ordered = TRUE)) %>%
  arrange(variable)
# making table for CV
kwl_p_cvs_table_pre_post <-
  kwl_p_cvs_and_pvalues_pre_post %>%
  select(-data, -significant) %>%
  pivot_wider(names_from = period, values_from = cvs) %>%
  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
#---------------------------CV test for post-e and ch-con pair
kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
  kwl_p_metrics_long_post_chi %>%
  nest(-variable) %>%  # or nest(period, value) %>%
  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
                               ~bind_cols(asymptotic_test (
                                 .x$value,
                                 .x$period)))) %>%
  unnest(asymptotic_test)
# compute CV
kwl_p_cvs_pre_post <-
  kwl_p_metrics_long_post_chi %>%
  group_by(period, variable) %>%
  summarise(cvs = raster::cv(value, na.rm = TRUE))
# join p-values to CVs
kwl_p_cvs_and_pvalues_post_chi <-
  kwl_p_cvs_pre_post %>%
  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
  mutate(variable = factor(variable, ordered = TRUE)) %>%
  arrange(variable)
# making table for CV
kwl_p_cvs_table_post_chi <-
  kwl_p_cvs_and_pvalues_post_chi %>%
  select(-data, -significant) %>%
  pivot_wider(names_from = period, values_from = cvs) %>%
  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
# join two table together
kwl_p_cvs_table_all <-
  kwl_p_cvs_table_pre_post %>% 
  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
  select(-`post-e.y`) %>% 
  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
  select(variable, `before European`, European, Chinese,
         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
knitr::kable(kwl_p_cvs_table_all,
             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996]. The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
